Statically Typed String Sanitation Inside a Python: Technical Report

### 2014-10-16 sanitation-psp14/tr/strings_chapter.tex
Proof reading.

\usepackage{ stmaryrd }
\usepackage{verbatimbox}
\input{../att-icfp14/macros-atlam}
\usepackage{alltt}
\renewcommand{\ttdefault}{txtt}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\definecolor{light-gray}{gray}{0.5}
\usepackage{listings}
\usepackage{wasysym}
    \makeatletter
\usepackage{enumitem}
\usepackage{enumerate}

@@

\newtheorem{tr}{TR Item}
\renewcommand{\thetr}{\Alph{tr}} % "letter-numbered" theorems
\newtheorem{trexam}[tr]{TR Example}
\newtheorem{trthm}[tr]{TR Theorem}
\newtheorem{trlem}[tr]{TR Lemma}
\newtheorem{trdef}[tr]{TR Definition}

@@

\maketitle
\section{String and Language Replacement}

@@

This section presents proofs of lemmas and theorems about the type systems
presented in \cite{fulton2014}. In addition, we provide some examples to help 
motivate and explain definitions.

@@

Theorems and lemmas appearing in \cite{fulton2014} are numbered, while
supporting facts are lettered.

@@

\begin{trexam}[All the heads \\
of all the tails \\
can be more than \\
one head and tail]\label{exam:headsandtails}
$r \not = \lhead{r} \cdot \ltail{r}$.
\end{trexam}
\begin{proof}
A simple counter-example is $ab + cd$. 
Note that $\lhead{ab + cd} = a+c$ and $\ltail{ab + cd} = b + d$. 
Therefore, $\{ad,bc\} \subset \lang{\lhead{ab + cd} \cdot \ltail{ab + cd}}$ even though neither of these is in $\lang{r}$.
\end{proof}

\todo{clean up these paragraphs}
Example~\ref{exam:headsandtails} does not cause type soundness problems for ${\sf strcase}$ because $s \in \lang{r} \implies s \in \lhead{r}\cdot\ltail{r}$ is the
property required for soundness. Still, in a production implementation, it will make sense to massage the definitions of $\lhead{r}$ and $\ltail{r}$ so that type information is not unnecessarily lost during sub-string operations.

This is a general pattern in string operations -- $\lambdas$ simulates common operations on strings within the type system.
If there is an operation for concatenating to strings, we define an operation for concatenating to regular expressions.
If there is an operation for peeling off the first (n) characters of a string, then we define an operation for peeling off
the first (n) characters of a regular expression.

It is important to note, however, that we need not \emph{exactly} simulate the action of string operations using regular expressions.
In the case of concatenation, we lose some information -- of course the string ad is never in $r$, but after decomposing the string, our
type tells us this might actually be the case. That's okay, because the types are conservative in their approximation. Soundness is never violated.

In the case of string replacement, there are \emph{trivial} definitions of substitution (on strings) and replacement (on languages) which over-approximate 
the effect of a substitution. Closing these gaps in approximation is important, and motivates the string operations portion of this TR.

@@

\begin{defn}[$\tt{subst}$]
We consider several choices in the previous section.
\end{defn}

\begin{defn}[$\tt{lreplace}$]
We consider several choices in the previous section.
\end{defn}

\begin{prop}[Closure] \label{thm:total}
  If $\lang{r}, \lang{r_1}$ and $\lang{r_2}$ are regular languages, then $\lang{\lreplace{r}{r_1}{r_2}}$ is also a regular language.
\end{prop}
\begin{proof}
Correctness implies closure.
\end{proof}

@@

In the S-T-Concat case, note that \ref{thm:total} implies that if $r_1$ and $r_2$ are regular expressions then so is $r_1 \cdot r_2$.

@@

To prove type safety and the security theorem for the big step semantics, we
first prove type safety for a small step semantics in Figure~\ref{sdlambdas}
and then extend this to the big step semantics in Figure~\ref{dlambdas} by
proving a correspondence between the semantics.

@@

Note that small step evaluation for $\lambdas$ is terminating:
%\begin{trthm}[Termination of small step semantics]
if $\emptyset \vdash e : \sigma$ then $e \mapsto^* v$ such that $v \ \val$.
%\end{trthm}
%\begin{proof}
We do not develop the full proof here, but note that the simply typed lambda calculus terminates.
For the string fragment, observe that the S-T- rules do not add any non-trivial binding structure because substitutions
$[e/x]e'$ may only occur in the special case where $e = \rstr{s}$, so that the length of the term
never increases and the number of free variables strictly decreases. Therefore, the standard normalization
argument proceeds without complication after fixing an evaluation order for the compatibility rules 
(all our other proofs are agnostic to evaluation order).
%\end{proof}

@@

\begin{trlem}[Progress of small step semantics.]
If $\vdash e : \sigma$ either $e \ \val$ or $e \mapsto e'$ for some $e'$.
\end{trlem}
\begin{proof}
The proof proceeds by induction on the derivation of $\vdash e : \sigma$.
\begin{itemize}[label=$ $,itemsep=1ex]
\item \textbf{$\lambda$ fragment}. Cases SS-T-Var, SS-T-Abs, and SS-T-App are exactly as in a proof of type safety for
the simply typed lambda calculus.

\item \textbf{S-T-Stringin-I}. Suppose $\vdash \rstr{s} : \stringin{s}$.
The $\rstr{s} \ \val$ by SS-E-RStr.

\item \textbf{S-T-Concat}. Suppose $\vdash \rconcat{e_1}{e_2} : \stringin{s}$.
By inversion and induction, $e_1 \mapsto e_1'$ or $e_1 \ \val$ and similarly
for $e_2$. 
If $e_1$ steps, then SS-E-Concat-Left applies and so
$\rconcat{e_1}{e_2} \mapsto \rconcat{e_1'}{e_2}$. Similarly, if $e_2$ steps
then $e$ steps by SS-E-Concat-Right.

In the remaining case, $e_1 \ \val$ and $e_2 \ \val$.
But then it follows by canonical forms that $e_1 = \rstr{s_1}$
and $e_2 = \rstr{s_2}$.
Note that $\rconcat{\rstr{s_1}}{\rstr{s_2}} \mapsto \rstr{s_1s_2}$ by SS-E-Concat.

\def\mline#1{\par\hspace*{-\leftmargin}\parbox{\textwidth}{\[#1\]}}
\item \textbf{S-T-Case}.
Suppose $e = \strcase{e_1}{e_2}{x,y.e_3}$.
By inversion, $e_1 : \stringin{r}$.
%\begin{align}
%&\vdash e_1 : \stringin{r} \\
%&\vdash e_2 : \sigma \\
%x:\stringin{\lhead{r}}, y : \stringin{\ltail{r}} &\vdash e_3 : \sigma
%\end{align}
%Actually this isn't even necessary...
%By (2), $e_2 \mapsto e_2'$ or $e_2 \val$. 
%In the former case, $e \mapsto \strcase{e_1}{e_2'}{x,y.e_3}$.
From this fact, induction, and canonical it follows that $e_1 \mapsto e_1'$ or $e_1 = \rstr{s}$.
In the former case, $e$ steps by S-E-Case-Left.
In the latter case, note that $s = \epsilon$ or $s = at$ for some string $t$.
If $s = \epsilon$ then $e$ steps by S-E-Case-$\epsilon$-Val,
and if $s = at$ the $e$ steps by S-E-Case-Concat.

\item \textbf{S-T-Replace}.
Suppose $e = \rreplace{r}{e_1}{e_2}$ and $e : \stringin{r'}$ where, by inversion of S-T-Replace,
\begin{align}
&\vdash e_1 : \stringin{r_1} \\
&\vdash e_2 : \stringin{r_2} \\
&\lreplace{r}{r_1}{r_2} = r'
\end{align}

By (1), inversion and induction $e_1 \ \val$ or $e_1 \mapsto e_1'$ for some $e_1'$,
If $e_1 \mapsto e_1'$ then $e$ steps by SS-E-Replace-Left.
Similarly, if $e_2$ steps then $e$ steps by SS-E-Replace-Right.
The only remaining case is where $e_1 \ \val$ and also $e_2 \ \val$.
But then by canonical forms, $e_1 = \strin{s_1}$ and $e_2 = \strin{s_2}$.
In this case, $e \mapsto \rstr{\subst{r}}{s_1}{s_2}$ by SS-E-Replace.

\item \textbf{S-T-SafeCoerce}.
Suppose that $\vdash \rcoerce{r}{e_1} : \stringin{r}$.
By inversion of S-T-SafeCoerce,
%\begin{align}
$\vdash e_1 : \stringin{r'}$% \\
%&\lang{r'} \subseteq \lang{r}
%\end{align}
By induction, $e_1 \ \val$ or $e_1 \mapsto e_1'$ for some $e_1'$.
If $e_1 \mapsto e_1'$ then $e$ steps by SS-E-SafeCoerce-Step.
Otherwise, $e_1 \ \val$ and by canonical forms $e_1 = \rstr{s}$.
In this case, $e = \rcoerce{r}{\rstr{s}} \mapsto \rstr{s}$ by SS-E-SafeCoerce.

\item \textbf{S-T-SafeCheck}
Suppose that $\vdash \rcheck{r}{e_1} : \stringin{r}$.
By inversion of S-T-Check:
\begin{align}
&\vdash e_0 : \stringin{r_0}  \\
x:\stringin{r} &\vdash e_1 : \sigma \\
&\vdash e_2 : \sigma \label{align6}
\end{align}

By (\ref{align6}) and induction,  $e_0 \mapsto e_0'$ or $e_0 \ \val$.
In the former case $e$ steps by SS-E-Check-StepRight.
Otherwise, by canonical forms $e_0 = \rstr{s}$.
By Lemma~\ref{thm:regexprops}, $s \in \lang{r_0}$ or else not.
In either case, the SS-E-Check-Ok and SS-E-Check-NotOk steps $e$.

@@

\item \textbf{SS-E-Concat-Left}.
Suppose $e = \rconcat{e_1}{e_1} \mapsto \rconcat{e_1'}{e_2}$.
By inversion of S-T-Concat, $\vdash e_1 : \stringin{r_1}$ where $\vdash e : \stringin{r_1r_2}$.
By induction if $e_1 \mapsto e_1'$ then $\vdash e_1' : \stringin{r_1}$.
Therefore, $\vdash \rconcat{e_1'}{e_2} : \stringin{r_1r_2}$.

\item {\textbf{SS-E-Concat-Right}}. Similar to SS-E-Concat-Left.

\item \textbf{SS-E-Concat}. Suppose $\rconcat{\rstr{s_1}}{\rstr{s_2}} : \stringin{r_1r_2}$ and
$\rconcat{\rstr{s_1}}{\rstr{s_2}} \mapsto \rstr{s_1s_2}$.
Then by inversion $\rstr{s_1} : \stringin{r_1}$ and similarly for $\rstr{s_2}$.
Therefore, $s_1 \in \lang{r_1}$ and $s_2 \in  \lang{r_2}$ from which it follows by Lemma~\ref{thm:regexprops}
that $s_1s_2 \in \lang{r_1r_2}$. Therefore, $\vdash \rstr{s_1s_2} : \stringin{r_1r_2}$ by S-T-Rstr.

\item \textbf{S-E-Case-Left}.
Suppose that $e = \strcase{e_1}{e_2}{x,y.e_3} \mapsto \strcase{e_1'}{e_2}{x,y.e_3}$
and $\emptyset \vdash e : \stringin{r}$.
By inversion of S-T-Case:
\begin{align}
&\vdash e_1 : \stringin{r} \label{align7} \\
&\vdash e_2 : \sigma \label{align8} \\ 
x:\stringin{\lhead{r}}, y : \stringin{\ltail{r}} &\vdash e_3 : \sigma \label{align9}
\end{align}
By (\ref{align7}) and the assumption that $e_1 \mapsto e_1'$, it follows by induction that $e_1' : \stringin{r}$.
This fact together with (\ref{align8}) and (\ref{align9}) implies by S-T-Case that
$\strcase{e_1'}{e_2}{x,y.e_3} : \sigma$.

\item \textbf{SS-E-Case-Right}. 
Suppose that $e = \strcase{e_1}{e_2}{x,y.e_3} \mapsto \strcase{e_1}{e_2'}{x,y.e_3}$
and $\emptyset \vdash e : \stringin{r}$.
By inversion of S-T-Case:
\begin{align}
&\vdash e_1 : \stringin{r} \label{align10}\\
&\vdash e_2 : \sigma \label{align11} \\ 
x:\stringin{\lhead{r}}, y : \stringin{\ltail{r}} &\vdash e_3 : \sigma \label{align12}
\end{align}
By (\ref{align11}) and the assumption that $e_2 \mapsto e_2'$, it follows by induction that $e_2' : \stringin{r}$.
This fact together with (\ref{align10}) and (\ref{align12}) implies by S-T-Case that 
$\strcase{e_1}{e_2'}{x,y.e_3} : \sigma$.

\item \textbf{SS-E-Case-$\epsilon$-Val}.
Suppose $e = \strcase{-}{e_2}{-}: \sigma$ and $e \mapsto e_2$.
By inversion of S-T-Case, $e_2 : \sigma$.

\item \textbf{SS-E-Case-Concat}.
Suppose that $e = \strcase{\rstr{as}}{e_2}{x,y.e_3} \mapsto [\rstr{a}, \rstr{s} / x,y] e_3$
and that $e : \sigma$.
By inversion of S-T-Case:
\begin{align}
&\vdash \rstr{as} : \stringin{r} \label{align13}\\
&\vdash \rstr{e_2} : \sigma \label{align14} \\ 
x:\stringin{\lhead{r}}, y : \stringin{\ltail{r}} &\vdash e_3 : \sigma \label{align15}
\end{align}

We know that $as \in \lang{r}$ by (\ref{align13}) and inversion of S-T-Rstr.
Therefore, $a \in \lhead{r}$ by definition of ${\sf lhead}$.
Furthermore, $\ltail{r} = ... | \delta_a{r} | ...$ by definition of ${\sf ltail}$.
Note that $s \in \lang{\delta_a{r}}$ by definition of the derivative, and so
$s \in \ltail{r}$

From these facts about $a$ and $s$ we know by S-T-Rstr that $\vdash \rstr{a} : \stringin{\lhead{r}}$
and $\vdash \rstr{s} : \stringin{\lhead{r}}$.
It follows by (\ref{align15}) that  $\vdash [\rstr{a}, \rstr{s} / x,y] e_3 : \sigma$.

@@

\item \textbf{Case SS-E-SafeCoerce}.
Suppose that $\rcoerce{r}{s_1} \mapsto \rstr{s_1}$ and that $\emptyset\vdash \rcoerce{r}{s_1} : \stringin{r}$.
By inversion of S-T-SafeCoerce we know that $s \in \lang{r}$ from which it follows by S-T-Rstr that $\emptyset\vdash s : \stringin{r}$.

\item \textbf{Case SS-E-Check-Ok}.
Suppose that $\rcheck{r}{\rstr{s}}{x.e_1}{e_2} \mapsto \subst{\rstr{s}}{x}{e_1}$,
$s \in \lang{r}$ and that $\emptyset \vdash \rcheck{r}{\rstr{s}}{x.e_1}{e_2} : \sigma$.
By inversion of S-T-Check, $x : \stringin{r} \vdash e_1 : \sigma$.
Note that $s \in \lang{r}$ implies that $s : \stringin{r}$ by S-T-RStr.
Therefore, $\emptyset \vdash \subst{\rstr{s}}{x}{e_1} : \sigma$

\item \textbf{Case SS-E-Check-NotOk}.
Suppose that $\rcheck{r}{\rstr{s}}{x.e_1}{e_2} \mapsto e_2$,
$s \not \in \lang{r}$ and that $\emptyset \vdash \rcheck{r}{\rstr{s}}{x.e_1}{e_2} : \sigma$.
By inversion of S-T-Check, $\emptyset \vdash e_2 : \sigma$.

@@

\begin{trthm}[Type Safety for small step semantics.] \label{thm:sstypesafte}
If $\emptyset\vdash e : \sigma$ then either $e \ \val$ or $e \mapsto^* e'$ and $\emptyset\vdash e' : \sigma$.
\end{trthm}
\begin{proof}
Follows directly from progress and preservation.
\end{proof}

@@

Before extending these the type safety result fir the small step semantics to the big step semantics, we first establish a correspondence between
the big step semantics in Figure~\ref{fig:sdlambdas} and the small step semantics in Figure~\ref{fig:dlambdas}.
We prove the relevant theorems for the $\mapsto$ relation because these proofs are easier. We then extend this result via the correspondence theorem
to the more concise big step semantics presented in \cite{fulton2014}.

@@

From these two facts we know via TR-Step that $\rreplace{r}{e_1}{e_2} \mapsto^* \rreplace{r}{e_1}{e_2}$.
Recall that $\lsubst{r}{s_1}{s_2} = s$, from which the conclusion follows.

@@

Establishing the other direction requires a minor lemma about the relationship between values and reflexivity in the big step semantics.

\begin{trthm}[Small Step Values are Reflexive in the Big Step Semantics]
If $v \ \val$ in the small step semantics then $v \sreduces v$.
\end{trthm}
\begin{proof}
If $v \ \val$ is derived from SS-E-RStr then S-E-RStr establishes the result.
Otherwise $v \ \val$ is derived from SS-E-Abs and so S-E-Abs establishes the result.
\end{proof}

\begin{trthm}[Semantic Correspondence for $\lambdas$ (Part II)]
If $e \vdash \tau$, $e \mapsto^* v$ and $v \ \val$ then $e \sreduces v$.  
\end{trthm}
\begin{proof}
The proof proceeds by structural induction on $e$.

@@

\end{mathpar}
\caption{Small step semantics for $\lambdas$. Extends \ref{fig:lambda}.}
\label{fig:dslambdas}
\end{figure}

>>>>

\usepackage{ stmaryrd }
\usepackage{verbatimbox}
\input{../../att-icfp14/macros-atlam}
\usepackage{alltt}
\renewcommand{\ttdefault}{txtt}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\definecolor{light-gray}{gray}{0.5}
\usepackage{listings}
\usepackage{wasysym}
    \makeatletter
\usepackage{enumitem}
\usepackage{enumerate}

@@

\newtheorem{tr}{TR Item}
\renewcommand{\thetr}{\Alph{tr}} % "letter-numbered" theorems
\newtheorem{trexam}[tr]{TR Example}
\newtheorem{trthm}[tr]{TR Theorem}
\newtheorem{trlem}[tr]{TR Lemma}
\newtheorem{trdef}[tr]{TR Definition}
\newtheorem{trconj}[tr]{TR Conjecture}

@@

\maketitle

\begin{abstract}
This Technical Report contains supporting materials for Statically Typed
String Sanitation Inside a Python \cite{fulton2014} including proofs of lemmas
and theorems asserted in the paper, examples, and additional discussion of
the paper's technical content.
\end{abstract}

@@

This section presents proofs of lemmas and theorems about the type systems
presented in \cite{fulton2014}, the accompanying paper. In addition, we provide some examples to help 
motivate and explain definitions.

@@

Theorems and lemmas appearing in \cite{fulton2014} are numbered, while
supporting facts appearing only in the Technical Report are lettered.

@@

\begin{trexam}[All the heads of all the tails can be more than one head and tail]\label{exam:headsandtails}
$r \not = \lhead{r} \cdot \ltail{r}$.
\end{trexam}
\begin{proof}
A simple counter-example is $ab + cd$. 
Note that $\lhead{ab + cd} = a+c$ and $\ltail{ab + cd} = b + d$. 
Therefore, $\{ad,bc\} \subset \lang{\lhead{ab + cd} \cdot \ltail{ab + cd}}$ even though neither of these is in $\lang{r}$.
\end{proof}

Example~\ref{exam:headsandtails} does not imply a counter-example to type soundness because $s \in \lang{r} \implies s \in \lhead{r}\cdot\ltail{r}$ is the
property required for soundness. Still, in a production implementation, it will make sense to massage the definitions of $\lhead{r}$ and $\ltail{r}$ so that type information is not unnecessarily lost during substring operations.

This is a general pattern in string operations: $\lambdas$ simulates -- within the type system -- common operations on strings. 
If there is an operation for concatenating to strings, we define an operation for concatenating two regular expressions.
If there is an operation for peeling off the first (n) characters of a string, then we define an operation for 
converting a regular expression $r$ into a regular expression $r'$ which recognizes any $n^{th}$ suffix of a string in $r$.

It is important to note, however, that the type system need not \emph{exactly} simulate the action of string operations.
In the case of concatenation, we lose some information because more string values are possible -- according the types --
than are actually possible in the dynamic semantics.
Soundness is not lost because the types are conservative in their approximation.

In the case of string replacement, there are \emph{trivial} definitions of substitution (on strings) and replacement (on languages) which over-approximate 
the effect of a substitution. Closing these gaps in approximation is important, and motivates the string operations portion of this technical report.

@@

\begin{defn}[$\tt{subst}$]
We consider several choices in the string operations section.
\end{defn}

\begin{defn}[$\tt{lreplace}$]
We consider several choices in the string operations section.
\end{defn}

\begin{prop}[Closure] \label{thm:total}
  If $\lang{r}, \lang{r_1}$ and $\lang{r_2}$ are regular languages, then $\lang{\lreplace{r}{r_1}{r_2}}$ is also a regular language.
\end{prop}
\begin{proof}
This result is proven for various formulations in the next section.
\end{proof}

@@

In the S-T-Concat case, note that Lemma~\ref{thm:total} implies that if $r_1$ and $r_2$ are regular expressions then so is $r_1 \cdot r_2$.

@@

To prove type safety and the security theorems for the big step semantics, we
first prove type safety for a small step semantics in Figure~\ref{fig:sdlambdas}
and then extend this to the big step semantics in Figure~\ref{fig:dlambdas} by
proving a correspondence between the semantics.

@@

\begin{trconj}
if $\emptyset \vdash e : \sigma$ then $e \mapsto^* v$ such that $v \ \val$.
\end{trconj}
%\begin{trthm}[Termination of small step semantics]
%\end{trthm}
%\begin{proof}
We do not develop the full proof here, but note that the simply typed lambda calculus terminates.
For the string fragment, observe that the S-T- rules do not add any non-trivial binding structure because substitutions
$[e/x]e'$ may only occur in the special case where $e = \rstr{s}$, so that the length of the term
never increases and the number of free variables strictly decreases. Therefore, the standard normalization
argument proceeds without complication after fixing an evaluation order for the compatibility rules 
(all our other proofs are agnostic to evaluation order).
%\end{proof}

@@

\begin{trlem}[Progress of small step semantics.]
If $\emptyset\vdash e : \sigma$ either $e \ \val$ or $e \mapsto e'$ for some $e'$.
\end{trlem}
\begin{proof}
The proof proceeds by induction on the derivation of $\emptyset\vdash e : \sigma$.
\begin{itemize}[label=$ $,itemsep=1ex]
\item \textbf{$\lambda$ fragment}. Cases SS-T-Var, SS-T-Abs, and SS-T-App are exactly as in a proof of type safety for
the simply typed lambda calculus.

\item \textbf{S-T-Stringin-I}. Suppose $\emptyset\vdash \rstr{s} : \stringin{s}$.
The $\rstr{s} \ \val$ by SS-E-RStr.

\item \textbf{S-T-Concat}. Suppose $\emptyset\vdash \rconcat{e_1}{e_2} : \stringin{s}$.
By inversion and induction, $e_1 \mapsto e_1'$ or $e_1 \ \val$ and similarly
for $e_2$. 
If $e_1$ steps, then SS-E-Concat-Left applies and so
$\rconcat{e_1}{e_2} \mapsto \rconcat{e_1'}{e_2}$. Similarly, if $e_2$ steps
then $e$ steps by SS-E-Concat-Right.

In the remaining case, $e_1 \ \val$ and $e_2 \ \val$.
But then it follows by Canonical Forms that $e_1 = \rstr{s_1}$
and $e_2 = \rstr{s_2}$.
Finally, by SS-E-Concat, \\ $\rconcat{\rstr{s_1}}{\rstr{s_2}} \mapsto \rstr{s_1s_2}$.

\def\mline#1{\par\hspace*{-\leftmargin}\parbox{\textwidth}{\[#1\]}}
\item \textbf{S-T-Case}.
Suppose $e = \strcase{e_1}{e_2}{x,y.e_3}$. 
By inversion,\\ $\emptyset\vdash e_1 : \stringin{r}$.
%\begin{align}
%&\vdash e_1 : \stringin{r} \\
%&\vdash e_2 : \sigma \\
%x:\stringin{\lhead{r}}, y : \stringin{\ltail{r}} &\vdash e_3 : \sigma
%\end{align}
%Actually this isn't even necessary...
%By (2), $e_2 \mapsto e_2'$ or $e_2 \val$. 
%In the former case, $e \mapsto \strcase{e_1}{e_2'}{x,y.e_3}$.
From this fact, induction, and Canonical Forms it follows that $e_1 \mapsto e_1'$ or $e_1 = \rstr{s}$.
In the former case, $e$ steps by S-E-Case-Left.
In the latter case, note that $s = \epsilon$ or $s = at$ for some string $t$.
If $s = \epsilon$ then $e$ steps by S-E-Case-$\epsilon$-Val,
and if $s = at$ the $e$ steps by S-E-Case-Concat.

\item \textbf{S-T-Replace}.
Suppose $e = \rreplace{r}{e_1}{e_2}$ and $\emptyset\vdash e : \stringin{r'}$ where, by inversion of S-T-Replace,
\begin{align}
&\emptyset\vdash e_1 : \stringin{r_1} \\
&\emptyset\vdash e_2 : \stringin{r_2} \\
&\lreplace{r}{r_1}{r_2} = r'
\end{align}

By (1), inversion and induction $e_1 \ \val$ or $e_1 \mapsto e_1'$ for some $e_1'$,
If $e_1 \mapsto e_1'$ then $e$ steps by SS-E-Replace-Left.
Similarly, if $e_2$ steps then $e$ steps by SS-E-Replace-Right.
The only remaining case is where $e_1 \ \val$ and also $e_2 \ \val$.
But then by Canonical Forms, $e_1 = \strin{s_1}$ and $e_2 = \strin{s_2}$.
Therefore, $e \mapsto \rstr{\lsubst{r}{s_1}{s_2}}$ by SS-E-Replace.

\item \textbf{S-T-SafeCoerce}.
Suppose that $\emptyset\vdash \rcoerce{r}{e_1} : \stringin{r}$.
By inversion of S-T-SafeCoerce,
%\begin{align}
$\emptyset\vdash e_1 : \stringin{r'}$ for $\lang{r'} \subseteq \lang{r}$.% \\
%&\lang{r'} \subseteq \lang{r}
%\end{align}
By induction, $e_1 \ \val$ or $e_1 \mapsto e_1'$ for some $e_1'$.
If $e_1 \mapsto e_1'$ then $e$ steps by SS-E-SafeCoerce-Step.
Otherwise, $e_1 \ \val$ and by Canonical Forms $e_1 = \rstr{s}$.
In this case, $e = \rcoerce{r}{\rstr{s}} \mapsto \rstr{s}$ by SS-E-SafeCoerce.

\item \textbf{S-T-SafeCheck}
Suppose that $\emptyset\vdash \rcheck{r}{e_0}{x.e_1}{e_2} : \stringin{r}$.
By inversion of S-T-Check:
\begin{align}
&\vdash e_0 : \stringin{r_0}  \\
x:\stringin{r} &\vdash e_1 : \sigma \\
&\vdash e_2 : \sigma \label{align6}
\end{align}

By (\ref{align6}) and induction,  $e_0 \mapsto e_0'$ or $e_0 \ \val$.
In the former case $e$ steps by SS-E-Check-StepRight.
Otherwise, $e_0 = \rstr{s}$ by Canonical Forms.
By Lemma~\ref{thm:regexprops}, either $s \in \lang{r_0}$ or $s \not\in\lang{r_0}$.
In the former case $e$ takes a step by SS-E-Check-Ok.
In the latter case $e$ takes a step by SS-E-Check-NotOk.

@@

\item \textbf{SS-E-Concat-Left}.
Suppose $e = \rconcat{e_1}{e_2} \mapsto \rconcat{e_1'}{e_2}$ and $e_1 \mapsto e_1'$.
By inversion of S-T-Concat, $\emptyset\vdash e_1 : \stringin{r_1}$ where \\$\emptyset\vdash e : \stringin{r_1r_2}$.
By induction, if $e_1 \mapsto e_1'$ then $\emptyset\vdash e_1' : \stringin{r_1}$.
Therefore, $\emptyset\vdash \rconcat{e_1'}{e_2} : \stringin{r_1r_2}$.

\item {\textbf{SS-E-Concat-Right}}. \todo{Similar to SS-E-Concat-Left.}

\item \textbf{SS-E-Concat}. Suppose $\emptyset\vdash\rconcat{\rstr{s_1}}{\rstr{s_2}} : \stringin{r_1r_2}$ and
$\rconcat{\rstr{s_1}}{\rstr{s_2}} \mapsto \rstr{s_1s_2}$.
Then by inversion $\emptyset\vdash\rstr{s_1} : \stringin{r_1}$ and similarly for $\rstr{s_2}$.
Therefore, $s_1 \in \lang{r_1}$ and\\ $s_2 \in  \lang{r_2}$ from which it follows by Lemma~\ref{thm:regexprops}
that $s_1s_2 \in \lang{r_1r_2}$. Therefore, $\emptyset\vdash \rstr{s_1s_2} : \stringin{r_1r_2}$ by S-T-Rstr.

\item \textbf{S-E-Case-Left}.
Suppose that $e = \strcase{e_1}{e_2}{x,y.e_3}$ and also that
$e \mapsto \strcase{e_1'}{e_2}{x,y.e_3}$
and $\emptyset \vdash e : \stringin{r}$.
By inversion of S-T-Case:
\begin{align}
\emptyset&\vdash e_1 : \stringin{r} \label{align7} \\
\emptyset&\vdash e_2 : \sigma \label{align8} \\ 
x:\stringin{\lhead{r}}, y : \stringin{\ltail{r}} &\vdash e_3 : \sigma \label{align9}
\end{align}
By (\ref{align7}) and the assumption that $e_1 \mapsto e_1'$, it follows by induction that $\emptyset\vdash e_1' : \stringin{r}$.
This fact together with (\ref{align8}) and (\ref{align9}) implies by S-T-Case that
$\emptyset\vdash\strcase{e_1'}{e_2}{x,y.e_3} : \sigma$.

\item \textbf{SS-E-Case-Right}. 
We have that $e = \strcase{e_1}{e_2}{x,y.e_3}$, Suppose $e \mapsto \strcase{e_1}{e_2'}{x,y.e_3}$
and $\emptyset \vdash e : \stringin{r}$.
By inversion of S-T-Case:
\begin{align}
\emptyset&\vdash e_1 : \stringin{r} \label{align10}\\
\emptyset&\vdash e_2 : \sigma \label{align11} \\ 
x:\stringin{\lhead{r}}, y : \stringin{\ltail{r}} &\vdash e_3 : \sigma \label{align12}
\end{align}
By (\ref{align11}) and the assumption that $e_2 \mapsto e_2'$, it follows by induction that $\emptyset\vdash e_2' : \stringin{r}$.
This fact together with (\ref{align10}) and (\ref{align12}) implies by S-T-Case that 
$\emptyset\vdash\strcase{e_1}{e_2'}{x,y.e_3} : \sigma$.

\item \textbf{SS-E-Case-$\epsilon$-Val}.
Suppose: \begin{align*}
&e = \strcase{-}{e_2}{-} \\
&\emptyset\vdash e: \sigma \\
&e \mapsto e_2
\end{align*}
By inversion of S-T-Case, $e_2 : \sigma$.

\item \textbf{SS-E-Case-Concat}.
Suppose that $e = \strcase{\rstr{as}}{e_2}{x,y.e_3} \mapsto [\rstr{a}, \rstr{s} / x,y] e_3$
and that $\emptyset\vdash e : \sigma$.
By inversion of S-T-Case:
\begin{align}
\emptyset&\vdash \rstr{as} : \stringin{r} \label{align13}\\
\emptyset&\vdash \rstr{e_2} : \sigma \label{align14} \\ 
x:\stringin{\lhead{r}}, y : \stringin{\ltail{r}} &\vdash e_3 : \sigma \label{align15}
\end{align}

We know that $as \in \lang{r}$ by (\ref{align13}) and inversion of S-T-Rstr.
Therefore, $a \in \lang{\lhead{r}}$ by definition of ${\sf lhead}$.
Furthermore, $\ltail{r} = ... | \delta_a{r} | ...$ by definition of ${\sf ltail}$.
Note that $s \in \lang{\delta_a{r}}$ by definition of the derivative, and so
$s \in \lang{\ltail{r}}$

From these facts about $a$ and $s$ we know by S-T-Rstr that $\emptyset\vdash \rstr{a} : \stringin{\lhead{r}}$
and $\emptyset\vdash \rstr{s} : \stringin{\lhead{r}}$.
It follows by (\ref{align15}) that  $\emptyset\vdash [\rstr{a}, \rstr{s} / x,y] e_3 : \sigma$.

@@

\item \textbf{Case SS-E-SafeCoerce}.
Suppose that $\rcoerce{r}{s_1} \mapsto \rstr{s_1}$ and that $\emptyset\vdash \rcoerce{r}{s_1} : \stringin{r}$.
By inversion of S-T-SafeCoerce we know that $s \in \lang{r}$. Therefore, $\emptyset\vdash s : \stringin{r}$.

\item \textbf{Case SS-E-Check-Ok}.
Suppose $\rcheck{r}{\rstr{s}}{x.e_1}{e_2} \mapsto \subst{\rstr{s}}{x}{e_1}$,
$s \in \lang{r}$, and $\emptyset \vdash \rcheck{r}{\rstr{s}}{x.e_1}{e_2} : \sigma$.
By inversion of S-T-Check, $x : \stringin{r} \vdash e_1 : \sigma$.
Note that $s \in \lang{r}$ implies that $s : \stringin{r}$ by S-T-RStr.
Therefore, $\emptyset \vdash \subst{\rstr{s}}{x}{e_1} : \sigma$.

\item \textbf{Case SS-E-Check-NotOk}.
Suppose $\rcheck{r}{\rstr{s}}{x.e_1}{e_2} \mapsto e_2$,
$s \not \in \lang{r}$, and $\emptyset \vdash \rcheck{r}{\rstr{s}}{x.e_1}{e_2} : \sigma$.
By inversion of S-T-Check, $\emptyset \vdash e_2 : \sigma$.

@@

\begin{trthm}[Type Safety for small step semantics.] \label{thm:sstypesafte}
If $\emptyset\vdash e : \sigma$ then either $e \ \val$ or $e \mapsto^* e'$ and $\emptyset\vdash e' : \sigma$.
\end{trthm}
\begin{proof}
\todo{Follows directly from progress and preservation.}
\end{proof}

@@

Before extending the previous theorem to the big step semantics, we first establish a correspondence between
the big step semantics in Figure~\ref{fig:sdlambdas} and the small step semantics in Figure~\ref{fig:dlambdas}.

@@

From these two facts we know via RT-Step that $\rreplace{r}{e_1}{e_2} \mapsto^* \rreplace{r}{e_1}{e_2}$.
Recall that $\lsubst{r}{s_1}{s_2} = s$, from which the conclusion follows.

@@

%Establishing the other direction requires a minor lemma about the relationship between values and reflexivity in the big step semantics.
%
%\begin{trthm}[Small Step Values are Reflexive in the Big Step Semantics]
%If $v \ \val$ in the small step semantics then $v \sreduces v$.
%\end{trthm}
%\begin{proof}
%If $v \ \val$ is derived from SS-E-RStr then S-E-RStr establishes the result.
%Otherwise $v \ \val$ is derived from SS-E-Abs and so S-E-Abs establishes the result.
%\end{proof}

\begin{trthm}[Semantic Correspondence for $\lambdas$ (Part II)]
If $\emptyset\vdash e : \sigma$, $e \mapsto^* v$ and $v \ \val$ then $e \sreduces v$.  
\end{trthm}
\begin{proof}
The proof proceeds by structural induction on $e$.

@@

\end{mathpar}
\caption{Small step semantics for $\lambdas$. Extends \ref{fig:lambda}.}
\label{fig:sdlambdas}
\end{figure}
